---
title: "Exploratory Correlation Analysis at Census Radius Level"
format: html
execute:
  echo: true
---


### Exploratory Correlation Analysis (Household Indicators)

Initial correlation analysis of the household and population census dataset to examine its internal structure and identify highly redundant variables.

### Correlation matrix (household indicators)

```{r}
library(dplyr)
library(ggplot2)

num_cols <- df_hogares_pct[2:40]
hogares_cor_mat <- cor(num_cols, use = "pairwise.complete.obs")

# Show dimensions
cat("Dimensiones de la matriz de correlación:", dim(hogares_cor_mat), "\n")
```

### Correlation matrix

```{r}
hogares_cor_mat[1:min(6, nrow(hogares_cor_mat)), 1:min(6, ncol(hogares_cor_mat))]
```

### Heatmap of correlation

```{r}
# Compute hierarchical clustering order so strongly correlated variables are plotted close together
ord <- hclust(as.dist(1 - abs(hogares_cor_mat)))$order

# Reorder variable names according to clustering
vars_ord <- rownames(hogares_cor_mat)[ord]

# Convert correlation matrix to long format for ggplot and set factor levels
cor_df <- as.data.frame(as.table(hogares_cor_mat)) |>
  mutate(
    var1 = factor(Var1, levels = vars_ord), 
    var2 = factor(Var2, levels = vars_ord)  
  )

# Draw heatmap
cor_df |> ggplot(aes(x = var1, y = var2, fill = Freq)) +
  geom_raster() +
  # Blue = negative correlation, white = 0, red = positive correlation
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_fixed() + 
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

### Pairs of variables with highest correlation

```{r}
cor_df <- as.data.frame(as.table(hogares_cor_mat))
colnames(cor_df) <- c("var1", "var2", "corr")
cor_df <- cor_df |> mutate(abs_corr = abs(corr))

cor_df |> 
  filter(var1 != var2) |> 
  arrange(desc(abs_corr)) |> 
  slice_head(n = 12)
```

### Key Insights (Household Indicators)

Strong variable pairs:
hogares_sin_caneria_de_agua_en_la_vivienda ↔ hogares_sin_sistema_de_boton_cadena_o_mochila_para_limpieza_del_inodoro: 1.00
hogares_de_personas_solas ↔ hogares_de_1_persona: 0.999
hogares_con_jefes_con_primaria_completa_o_secundaria_incompleto ↔ hogares_con_jefes_con_secundaria_completa_o_mas: -0.978
hogares_propietarios_de_la_vivienda ↔ hogares_inquilinos_de_la_vivienda: -0.959
hogares_con_computadora ↔ hogares_con_jefes_con_secundaria_completa_o_mas: 0.939
hogares_con_hacinamiento_mas_de_2_personas_por_cuarto ↔ hogares_con_hacinamiento_critico_mas_de_3_personas_por_cuarto: 0.919



# Exploratory Correlation Analysis (Population Indicators)

This section repeats the same correlation workflow using the person-level dataset `df_personas_pct`.

```{r}
library(dplyr)
library(ggplot2)

# Define the personal indicators correlation dataset from df_personas_pct using columns 2 through 60.
# Keep only the numeric columns in that range before computing correlations.
df_personas_pct_cor <- df_personas_pct[2:60] |> select(where(is.numeric))

personas_cor_mat <- cor(df_personas_pct_cor, use = "pairwise.complete.obs")

cat("Dimensions of the personal indicators correlation matrix:", dim(personas_cor_mat), "\n")
```

### Correlation matrix preview

```{r}
personas_cor_mat[1:min(6, nrow(personas_cor_mat)), 1:min(6, ncol(personas_cor_mat))]
```

### Correlation heatmap

```{r}
ord_personas <- hclust(as.dist(1 - abs(personas_cor_mat)))$order
vars_ord_personas <- rownames(personas_cor_mat)[ord_personas]
cor_df_personas <- as.data.frame(as.table(personas_cor_mat)) |>
  mutate(
    var1 = factor(Var1, levels = vars_ord_personas),
    var2 = factor(Var2, levels = vars_ord_personas)
  )

cor_df_personas |> ggplot(aes(x = var1, y = var2, fill = Freq)) +
  geom_raster() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_fixed() +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

### Top correlated population variable pairs

```{r}
cor_df_personas <- as.data.frame(as.table(personas_cor_mat))
colnames(cor_df_personas) <- c("var1", "var2", "corr")
cor_df_personas <- cor_df_personas |> mutate(abs_corr = abs(corr))

cor_df_personas |> 
  filter(var1 != var2) |> 
  arrange(desc(abs_corr)) |> 
  slice_head(n = 12)
```

### Key Insights (Population Indicators)

- The population-level data contains very strong relationships between these variable pairs:
  - `poblacion_de_18_anos_y_mas` ↔ `poblacion_de_0_a_17_anos`: -1.000
  - `total_de_mujeres_en_hogares_familiares` ↔ `total_de_varones_en_hogares_familiares`: -1.000
  - `personas_que_reciben_jubilacion_65_anos_y_mas` ↔ `personas_que_reciben_jubilacion_o_pension_65_anos_y_mas`: 0.9988
  - `poblacion_de_13_a_17_sin_secundaria_completa` ↔ `poblacion_de_13_a_17_anos`: 0.9971
  - `poblacion_de_6_a_12_que_asiste` ↔ `poblacion_de_6_a_12_anos`: 0.9961
  - `obra_social_o_prepaga_incluye_pami` ↔ `solo_salud_publica`: -0.9960


